Estimating duration in the presence of misclassification

### 2024-03-25 ATACCC.tex
proof and format numbers

\textcite{hakkiOnset} uses weakly informative priors, similar to \textcite{singanayagamCommunity}, see \cref{ATACCC:table:hakki-priors}.
I reran \textcite{hakkiOnset}'s analysis using R~4.3.1~\autocite{R-4-3-1}.
The inference was performed using RStan~2.32.3~\autocite{RStan-2-32-3} and Stan~2.32.2~\autocite{Stan-2-32-2}.
Inference settings were the same as \textcite{hakkiOnset}, which included 8 chains with a total of 8000 iterations each with the first 3000 used as warm-up.
\begin{table}
\begin{tabular}{l c l l}
    Parameter description & Symbol & Distribution \\
    \hline \\
    Mean max log viral load & $\mu_{v_\text{max}, g}$ & $\N(15, 15^2)$ \\
    Standard deviation of max log viral load & $\sigma_{v_\text{max}}$ & $\N_+(0, 10^2)$ \\
    Mean rate of log viral load increase & $\mu_{a, g}$ & $\N(1.25, 0.75^2)$ \\
    Standard deviation of rate of log viral load increase & $\sigma_{a}$ & $\N_+(0, 1)$ \\
    Mean rate of log viral load decrease & $\mu_{b, g}$ & $\N(0.5, 1.4^2)$ \\
    Standard deviation of rate of log viral load decrease & $\sigma_{b}$ & $\N_+(0, 1)$ \\
    Correlation of viral load parameters & $\matr{C}_\phi$ & $\LKJ(1)$ \\
    Negative log of rate of test failure & $-\log \rho$ & $\N(3, 0.5)$ \\
    Mean log viral load for test failures & $x_0$ & $\N(0, 3^2)$ \\
    Standard deviation of log viral load for test failures & $\sigma_0$ & $\N_+(3, 3^2)$ \\
    Standard deviation of log viral load observations & $\sigma_v$ & $1 + \N_+(2, 3^2)$ \\
    Time from first test to peak & $\tau_{\max,i}$ & $\N(5,5)$
\end{tabular}
\caption[Viral load model priors]{Priors for each parameter, taken from \textcite{hakkiOnset}. $\N_+$ indicates a normal distribution truncated to the non-negative reals. Details of the distributions, including their parameterizations, are in \cref{E-distributions}. The elements of $\vec{\mu}_{\phi,g}$ and $\vec{\sigma}_\phi$ have independent priors placed on them, as specified in the first lines of the table. Priors are the same for all $g$ and $i$. \label{ATACCC:table:hakki-priors}}
\end{table}

@@

\begin{algorithm}
\For{$i = 1, \dots, M$}{
  \For{$j = 1, \dots, M$}{
    Sample an individual-level random effects vector $\vec{\phi}_{i,j} \dist \LN(\vec{\mu}_{\phi,g}, \matr{\Sigma}_\phi)$ \;
    Calculate $d_{i,j}$ using \cref{ATACCC:eq:di} \;
  }
  \For{$d = 1, \dots, \dmax$}{
    Set $F_{D,g}^{(i)}(d) = \frac{1}{M} \sum_{j=1}^M \I(d_{i,j} \leq d + 0.5)$ \;
  }
}
  \caption{%
  Algorithm to calculate the Monte Carlo integral in \cref{ATACCC:eq:FDintegral}, producing the posterior distribution of $F_{D,g}(d)$.
  The posterior distribution is represented as a sample $F_{D,g}^{(1)}, \dots, F_{D,g}^{(N)}$.
  $\vec{\mu}_\phi^{(i)}$ and $\matr{\Sigma}_\phi^{(i)}$ denote the $i$th posterior sample of $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$ respectively, taken from the MCMC output.
  $M$ is the number of individuals to draw, I found $M = 100,000$ produced stable estimates of the tail of $F_{D,g}(d)$ if $\dmax = 45$.
 }
 \label{ATACCC:alg:MonteCarlo}
\end{algorithm}

>>>>

\textcite{hakkiOnset} uses weakly informative priors, similar to \textcite{singanayagamCommunity}, see \cref{ATACCC:table:hakki-priors}.
I reran \textcite{hakkiOnset}'s analysis using R~4.3.1~\autocite{R-4-3-1}.
The inference was performed using RStan~2.32.3~\autocite{RStan-2-32-3} and Stan~2.32.2~\autocite{Stan-2-32-2}.
Inference settings were the same as \textcite{hakkiOnset}, which included 8 chains with a total of \numprint{8000} iterations each with the first \numprint{3000} used as warm-up.
\begin{table}
\begin{tabular}{l c l l}
    Parameter description & Symbol & Distribution \\
    \hline \\
    Mean max log viral load & $\mu_{v_\text{max}, g}$ & $\N(15, 15^2)$ \\
    Standard deviation of max log viral load & $\sigma_{v_\text{max}}$ & $\N_+(0, 10^2)$ \\
    Mean rate of log viral load increase & $\mu_{a, g}$ & $\N(1.25, 0.75^2)$ \\
    Standard deviation of rate of log viral load increase & $\sigma_{a}$ & $\N_+(0, 1)$ \\
    Mean rate of log viral load decrease & $\mu_{b, g}$ & $\N(0.5, 1.4^2)$ \\
    Standard deviation of rate of log viral load decrease & $\sigma_{b}$ & $\N_+(0, 1)$ \\
    Correlation of viral load parameters & $\matr{C}_\phi$ & $\LKJ(1)$ \\
    Negative log of rate of test failure & $-\log \rho$ & $\N(3, 0.5)$ \\
    Mean log viral load for test failures & $x_0$ & $\N(0, 3^2)$ \\
    Standard deviation of log viral load for test failures & $\sigma_0$ & $\N_+(3, 3^2)$ \\
    Standard deviation of log viral load observations & $\sigma_v$ & $1 + \N_+(2, 3^2)$ \\
    Time from first test to peak & $\tau_{\max,i}$ & $\N(5,5)$
\end{tabular}
\caption[Viral load model priors]{Priors for each parameter, taken from \textcite{hakkiOnset}. $\N_+$ indicates a normal distribution truncated to the non-negative reals. Details of the distributions, including their parameterizations, are in \cref{E-distributions}. The elements of $\vec{\mu}_{\phi,g}$ and $\vec{\sigma}_\phi$ have independent priors placed on them, as specified in the first lines of the table. Priors are the same for all $g$ and $i$. \label{ATACCC:table:hakki-priors}}
\end{table}

@@

\begin{algorithm}
\For{$i = 1, \dots, M$}{
  \For{$j = 1, \dots, M$}{
    Sample an individual-level random effects vector $\vec{\phi}_{i,j} \dist \LN(\vec{\mu}_{\phi,g}, \matr{\Sigma}_\phi)$ \;
    Calculate $d_{i,j}$ using \cref{ATACCC:eq:di} \;
  }
  \For{$d = 1, \dots, \dmax$}{
    Set $F_{D,g}^{(i)}(d) = \frac{1}{M} \sum_{j=1}^M \I(d_{i,j} \leq d + 0.5)$ \;
  }
}
  \caption{%
  Algorithm to calculate the Monte Carlo integral in \cref{ATACCC:eq:FDintegral}, producing the posterior distribution of $F_{D,g}(d)$.
  The posterior distribution is represented as a sample $F_{D,g}^{(1)}, \dots, F_{D,g}^{(N)}$.
  $\vec{\mu}_\phi^{(i)}$ and $\matr{\Sigma}_\phi^{(i)}$ denote the $i$th posterior sample of $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$ respectively, taken from the MCMC output.
  $M$ is the number of individuals to draw, I found $M =\numprint{100000}$ produced stable estimates of the tail of $F_{D,g}(d)$ if $\dmax = 45$.
 }
 \label{ATACCC:alg:MonteCarlo}
\end{algorithm}

### 2024-03-25 biology-data.tex
proof and format numbers

False positives are very rare for SARS-CoV-2 RT-PCR tests.
In summer 2020, the CIS performed 208,730 PCR tests and found only 159 positives~\autocite[section 5]{cisMethodsONS}.
If these were all false positives, then the test specificity would be 99.9\%.
In reality, it is almost certain that some of these are true positives, and hence this is a lower-bound on the true test specificity.

@@

% Possible summaries
% \begin{enumerate}
%   \item \url{https://www.nature.com/articles/s41579-022-00822-w/figures/2}
%   \item \url{https://www.nature.com/articles/s41576-021-00360-w/figures/1}
%   \item \url{https://academic.oup.com/view-large/figure/305617142/ciaa1442_fig2.jpg}
% \end{enumerate}

% \section{Studies used ie this thesis} \label{biology-data:sec:studies}

% In this section, I expand on the major two studies I use in this thesis (ATACCC and CIS).
% Unlike many studies of SARS-CoV-2, these studies are not focused on specific groups such as healthcare workers or hospital patients.
% Therefore, the results from these studies are more generalisable to the entire UK population.

@@

The ATACCC study was a longitudinal study of contacts reported to the NHS Test and Trace contact tracing service~\autocite{hakkiOnset}.
Contacts were invited to participate if they were 5 years or older and  could be reached within 5 days of the index case's symptom onset date.
ATACCC enrolled a total of 738 contacts across two waves.
The first wave enrolled 393 contacts between 13th Sep 2020 and 31st Mar 2021, when pre-Alpha and Alpha variants were circulating.
The second wave enrolled 345 contacts between 24th May 2021 and 228th Oct 2021, when the Delta variant was circulating.
Contacts were swabbed daily for up to 20 days, with a variety of tests performed on the swabs; relevant for this thesis is the RT-PCR test for SARS-CoV-2.
For all positive test results, the viral load was measured using the Ct value and then converted to a viral load using a standard curve (see \cref{biology-data:sec:PCR-process}).
Other data such as demography and symptoms were also collected, although I do not use this information.
For further details see \textcite{singanayagamDuration,hakkiOnset}.

@@

The CIS~\autocite{CIS} was set up in April 2020 to provide gold-standard measurements of the prevalence of SARS-CoV-2 in the community.
It was a longitudinal study with a representative sample of private households, \ie excluding settings such as care homes and student halls.
The dataset is globally unique in providing a representative, longitudinal, and large-scale study across the pandemic.
Initially, it was limited to England, but expanded to cover the whole of the UK in September 2020.
Enrolment was continuous until 31st Jan 2022, with data collected until 13th Mar 2023~\autocite{weiRisk}.

@@

Up until July 2020, households were selected from those that had previously participated in an ONS survey~\autocite{CIStechData}.
Around 50\% of households agreed to participate, with around 90\% of eligible individuals (96,113 in total) in these households contributing multiple swabs to the survey.
After July 2020, addresses were also randomly sampled from a database of all UK addresses held by the ONS.
The response rate was lower, with only 12\% of households agreeing to participate, from which 86\% of eligible individuals (383,732 in total) contributed multiple swabs to the survey.

@@

In this thesis, I focus on the period from Mon 31st Aug 2020 until Sun 24th Jan 2021 (see \cref{E-intro:sec:aims}).
During the start of this period, the recruitment rate of CIS was expanded (see \cref{biology-data:fig:CIS-recruitment}(A)).
Following this, the recruitment then decreases and stabilises, as does the rate of swabs taken (see \cref{biology-data:fig:CIS-recruitment}(B)).
The recruitment pattern was chosen to hit targets regarding the fortnightly number of swabs.
Because testing becomes less frequent after the first 4 weeks, some continuous recruitment is needed to maintain the number of swabs, as well as to replace participants who drop out.

@@

Analyses requiring more fine-grained stratification (such as that in \cref{biology-data:sec:MRP}) or linking data from the same individual over time (such as that in \cref{E-perf-test,E-imperf-test}) must be performed within the SRS.
The SRS contains five 2.40\,GHz dual-core Intel Xeon E5-4650 processors and 164\,GB RAM.
These computational resources are shared between all users of the CIS SRS project and due consideration must be given to other users.
Processes which run for multiple days are liable to be cancelled.

>>>>

False positives are very rare for SARS-CoV-2 RT-PCR tests.
In summer 2020, the CIS performed\numprint{208730} PCR tests and found only 159 positives~\autocite[section 5]{cisMethodsONS}.
If these were all false positives, then the test specificity would be 99.9\%.
In reality, it is almost certain that some of these are true positives, and hence this is a lower-bound on the true test specificity.

@@

@@

The ATACCC study was a longitudinal study of contacts reported to the NHS Test and Trace contact tracing service~\autocite{hakkiOnset}.
Contacts were invited to participate if they were 5 years or older and  could be reached within 5 days of the index case's symptom onset date.
ATACCC enrolled a total of 738 contacts across two waves.
The first wave enrolled 393 contacts between 13 Sep 2020 and 31 Mar 2021, when pre-Alpha and Alpha variants were circulating.
The second wave enrolled 345 contacts between 24 May 2021 and 28 Oct 2021, when the Delta variant was circulating.
Contacts were swabbed daily for up to 20 days, with a variety of tests performed on the swabs; relevant for this thesis is the RT-PCR test for SARS-CoV-2.
For all positive test results, the viral load was measured using the Ct value and then converted to a viral load using a standard curve (see \cref{biology-data:sec:PCR-process}).
Other data such as demography and symptoms were also collected, although I do not use this information.
For further details see \textcite{singanayagamDuration,hakkiOnset}.

@@

The CIS~\autocite{CIS} was set up in April 2020 to provide gold-standard measurements of the prevalence of SARS-CoV-2 in the community.
It was a longitudinal study with a representative sample of private households, \ie excluding settings such as care homes and student halls.
The dataset is globally unique in providing a representative, longitudinal, and large-scale study across the pandemic.
Initially, it was limited to England, but expanded to cover the whole of the UK in September 2020.
Enrolment was continuous until 31 Jan 2022, with data collected until 13 Mar 2023~\autocite{weiRisk}.

@@

Up until July 2020, households were selected from those that had previously participated in an ONS survey~\autocite{CIStechData}.
Around 50\% of households agreed to participate, with around 90\% of eligible individuals (\numprint{96113} in total) in these households contributing multiple swabs to the survey.
After July 2020, addresses were also randomly sampled from a database of all UK addresses held by the ONS.
The response rate was lower, with only 12\% of households agreeing to participate, from which 86\% of eligible individuals (\numprint{383732} in total) contributed multiple swabs to the survey.

@@

In this thesis, I focus on the period from Mon 31 Aug 2020 until Sun 24 Jan 2021 (see \cref{E-intro:sec:aims}).
During the start of this period, the recruitment rate of CIS was expanded (see \cref{biology-data:fig:CIS-recruitment}(A)).
Following this, the recruitment then decreases and stabilises, as does the rate of swabs taken (see \cref{biology-data:fig:CIS-recruitment}(B)).
The recruitment pattern was chosen to hit targets regarding the fortnightly number of swabs.
Because testing becomes less frequent after the first 4 weeks, some continuous recruitment is needed to maintain the number of swabs, as well as to replace participants who drop out.

@@

Analyses requiring more fine-grained stratification (such as that in \cref{biology-data:sec:MRP}) or linking data from the same individual over time (such as that in \cref{E-perf-test,E-imperf-test}) must be performed within the SRS.
The SRS contains five \numprint[GHz]{2.40} dual-core Intel Xeon E5-4650 processors and \numprint[GB]{164} RAM.
These computational resources are shared between all users of the CIS SRS project and due consideration must be given to other users.
Processes which run for multiple days are liable to be cancelled.

### 2024-03-25 cis-imperfect-testing.tex
fix refs
proof and format numbers

% Discussion and interrogation of the data led me to
I hypothesised that the reason for the unexpectedly high number of single positive episodes (compared to simulation) was the presence of false negative results.
There are several strands of evidence supporting this hypothesis.
First, it is well known that RT-PCR testing can return false negatives  (see \cref{E-biology-data:sec:PCR}); I explicitly include them in the viral load model of \cref{E-ATACCC}.
Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
Furthermore, in \cref{imperf-test:sec:simulate}, I expanded the simulation study of \cref{E-perf-test:sec:simulation-study} to include false negatives (see  for more details).
This reproduced the CIS data more faithfully and exhibited the same issues when estimating the duration distribution (see \cref{imperf-test:fig:sim-single-pos}).
\begin{figure}
  \centering \includegraphics{cis-imperfect-testing/sim-single-positive-episodes}
  \caption[Single positive episodes in CIS simulation]{%
    Proportion of detected episodes which have a single positive episode.
    y-axis is the number of simulations, out of 1000, with the indicated proportion.
    Test sensitivity is $\psens$.
    The CIS data is shown as a vertical line.
  }
  \label{imperf-test:fig:sim-single-pos}
\end{figure}

@@

In this section I apply the approach described in this chapter to the CIS episodes dataset I described in \cref{E-perf-test:sec:approach}.
This is the 4,800 CIS infection episodes first detected between 10th Oct 2020 and 6th Dec 2020 inclusive and have negatives bounding the start and end time of the episode.

Unlike in the simulation studies, an uninformative prior on $\ntot$ led to implausible estimates of the duration distribution.
The uninformative prior led to high posterior estimates of $\ntot$, and hence an implausibly large number of episodes with durations of less than five days.
Therefore, I based an informative prior for $\ntot$, $\ntot \sim \NBc(\mu, r)$ (as in \cref{E-perf-test:sec:model}, see \cref{E-distributions} for definition), using pre-existing estimates of the total number of infections to give $\mu\inform$ and $r\inform$.
\Textcite{birrellRTM2} estimated the total number of infections in England over the time period I consider, with posterior mean \numprint{4136368} and standard deviation \numprint{27932}.
% This model gives a posterior mean of \numprint{4136368} cumulative infections in England in the time period I consider, with a posterior standard deviation of \numprint{27932}~\citePersonalComms{Paul Birrell}.
\todo{Insert correct citation for RTM paper 2 once available}
Approximating this distribution as a negative binomial and scaling the mean to the size of the CIS cohort gives the prior $\mu\inform = 25132$ and $r\inform = 22047)$.

>>>>

% Discussion and interrogation of the data led me to
I hypothesised that the reason for the unexpectedly high number of single positive episodes (compared to simulation) was the presence of false negative results.
There are several strands of evidence supporting this hypothesis.
First, it is well known that RT-PCR testing can return false negatives  (see \cref{E-biology-data:sec:PCR}); I explicitly include them in the viral load model of \cref{E-ATACCC}.
Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
Furthermore, in \cref{imperf-test:sec:simulate}, I expanded the simulation study of \cref{E-perf-test:sec:simulation-study} to include false negatives (see  for more details).
This reproduced the CIS data more faithfully and exhibited the same issues when estimating the duration distribution (see \cref{imperf-test:fig:sim-single-pos}).
\begin{figure}
  \centering \includegraphics{cis-imperfect-testing/sim-single-positive-episodes}
  \caption[Single positive episodes in CIS simulation]{%
    Proportion of detected episodes which have a single positive episode.
    y-axis is the number of simulations, out of \numprint{1000}, with the indicated proportion.
    Test sensitivity is $\psens$.
    The CIS data is shown as a vertical line.
  }
  \label{imperf-test:fig:sim-single-pos}
\end{figure}

@@

In this section I apply the approach described in this chapter to the CIS episodes dataset I described in \cref{perf-test:sec:data-notation}.
This is the \numprint{4800} CIS infection episodes first detected between 10 Oct 2020 and 6 Dec 2020 inclusive and have negatives bounding the start and end time of the episode.

Unlike in the simulation studies, an uninformative prior on $\ntot$ led to implausible estimates of the duration distribution.
The uninformative prior led to high posterior estimates of $\ntot$, and hence an implausibly large number of episodes with durations of less than five days.
Therefore, I based an informative prior for $\ntot$, $\ntot \sim \NBc(\mu, r)$ (as in \cref{E-perf-test:sec:model}, see \cref{E-distributions} for definition), using pre-existing estimates of the total number of infections to give $\mu\inform$ and $r\inform$.
\Textcite{birrellRTM2} estimated the total number of infections in England over the time period I consider, with posterior mean \numprint{4136368} and standard deviation \numprint{27932}.
% This model gives a posterior mean of \numprint{4136368} cumulative infections in England in the time period I consider, with a posterior standard deviation of \numprint{27932}~\citePersonalComms{Paul Birrell}.
\todo{Insert correct citation for RTM paper 2 once available}
Approximating this distribution as a negative binomial and scaling the mean to the size of the CIS cohort gives the prior $\mu\inform = \numprint{25132}$ and $r\inform = \numprint{22047}$.
